---
title: "matrix_eqtl_fcr"
author: "Mari Johnson"
date: "20/04/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = '~/GWAS_22/Fc_receptor/data')
```

### Aims
- Format PLINK data for input into MatrixEQTL R programme
- Format microarray gene expression data for mEQTL
- Run MatrixEQTL
- Get list of significant eQTLs in FCR region
- Eventually run for genome wide eQTL analysis, focus on 'top' snps
- Plot data for significant SNPs

### Possible associations to run

- Can either compare normalised expression values or Log2FC
- For now we are starting with correlating participant genotype info with D7 time point
- Day 7 see's the most pertubation of FcR gene expression, and is around time of diagnosis
- Will also look at day 1 expression 

```{bash Format Plink Data}

#Use all_enteric_QC data set

plink2 --bfile all_enteric_QC2 --recode oxford --out ox_enteric
plink --data ox_enteric --allow-no-sex --recode A-transpose --out traw_enteric
cat traw_enteric.traw | cut -f2,7- > geno_file
# Remove intermediate files:
rm -f *ped *hh *nosex *traw *map

```

```{r Format genotype data}
library(data.table)
library(dplyr)
sample <- fread("ox_enteric.csv")

write.table(sample, file="ox_enteric.sample", row.names = F, quote = F, sep = " ")
#geno_file has the data organised by snps (rows), participants (cols) - recoded 1-0
#Filter for fcyr genes

fcyr <- fread("~/GWAS_22/new_gwas/Post-imp/Merged/Assoc/T_fcy.txt")
geno <- fread("~/GWAS_22/Fc_receptor/data/geno_file")
fcyr <- fcyr[,3]

fc_geno <- left_join(fcyr, geno, by = "SNP")
#remove zero
names(fc_geno) <- gsub(pattern = "0_", replacement = "", x = names(fc_geno))

write.table(fc_geno, file="fcy.sample", row.names = F, quote = F, sep = " ")
#This is our geno file for matrix-eqtl 
```

```{r Format expression file}
#requires 


```